Robust Bayesian Analysis of Causal Inference Problems

### 2023-03-20 JASA-template.tex
just before meeting

\title{Robust Bayesian Analysis of Causal Inference Problems}
\author{Tathagata Basu$^1$}
\email{tathagatabasumaths@gmail.com}
\author{Matthias C.~M.~Troffaes$^2$}
\email{matthias.troffaes@durham.ac.uk}
\author{Jochen Einbeck$^{2,3}$}
\email{jochen.einbeck@durham.ac.uk}
\address{$^1$UMR CNRS 7253 Heudiasyc, Universit\'{e} de Technologie de
Compi\`{e}gne}
\address{$^2$Department of Mathematical Sciences, Durham University}
\address{$^3$Durham Research Methods Centre}

\begin{abstract}
Causal inference using observational data is an important aspect in
many fields such as epidemiology, social science, economics, etc. In
particular, our goal is to find the treatment effect on the subjects
along with the causal links between the variables and the outcome.
However, estimation for such problems are extremely 
difficult as the treatment effects may vary from subject 
to subject and modelling the underlying heterogeneity explicitly makes the 
problem practically unsolvable. Another issue we face is the 
dimensionality of the problem and we may wish to find a subset of 
explanatory variables. However, standard variable selection methods
tend to maximise the predictive performance of the outcome model only and
can also be sensitive with respect to the choice of priors, particularly 
in the case of limited information. So, in this paper, we suggest a 
robust Bayesian analysis of causal inferential methods for 
high-dimensional problems in a regressional framework. We consider
a set of spike and slab priors to obtain robust estimates for
both the treatment and outcome model. We are specifically interested 
in the importance of priors in the high dimensional causal inference
as well as the identifying the confounder variables. However, indicator
based confounder selection can be deceptive in some cases. Especially, 
when the predictor is strongly associated with either the treatment or 
the outcome. This increases the posterior expectation of the selection
indicators. To avoid that we also apply a post-hoc selection scheme
which successfully remove negligible non-zero effects from the model
attaining a sparser model. Finally, we illustrate
our result using synthetic \alert{and real} dataset.

@@

Bayesian approach in causal effect estimation has become a popular
topic in recent days. However, one of the earlier works on this can be
found in \cite{rubin1978}. Lately with the notion of high dimensional problems, Bayesian methodologies have become
more appealing. \citet{Crainiceanu2008} proposed a bi-level 
Bayesian model averaging based method for estimating the causal 
effect. \citet{wang2015} suggested BAC (or, Bayesian adjustment for
confounding) where they use an informative prior obtained from
the treatment model and apply them on the outcome model for
estimating causal effect. Several other methods were also
proposed to tackle confounders from the point of view of Bayesian
variable selection such as: \citet{Zigler2014}, \citet{Hahn2018} etc.

In this paper we take inspiration from the approach of \citet{koch2020}, where
they proposed a bilevel spike and slab prior for causal effect 
estimation. They considered a data-driven adaptive approach to
propose their prior which reduce the variance of the causal estimate. 
In our approach,
we suggest a sensitivity analysis based approach where instead
of using a single prior, we consider a set of priors \cite{BERGER1990303}. This is particularly interesting as in
many cases, causal effect estimation can be performed through a 
meta analysis and hence robust Bayesian analysis can be beneficial
\cite{raices_cruz22} under severe uncertainty.
Moreover, for some problems we have to rely on very limited data to
perform our Bayesian analysis and choosing a data-driven prior
can lead to overfitting. Therefore, as to propose our
robust Bayesian framework,
we consider a set of continuous spike and slab priors \cite{ishwaran2005}
for confounder selection and construct a Bayesian group LASSO
\cite{xu2015}. To perform the sensitivity analysis, we consider
a set of beta priors on the covariate selection probability
of the spike and slab priors. This sensitivity analysis allows us
to investigate the regression coefficients with respect to a different
level of sparsity of the model. Following that, we consider a post-hoc
variable selection method as suggested by \citet{hahn2015}. This
ensures that we can efficiently separate the confounder variables
with others and reduce the bias of treatment effect estimation.

The rest of the paper is organised as follows. In \cref{sec:causal}
we give a formal description of causal estimation problem in the
context of linear regression. \cref{sec:bayes} is focused on the
Bayesian analysis of causal inference problems, followed by the
motivation of a robust Bayesian analysis. In \cref{sec:sim}, we
provide our result of simulation studies under different 
scenarios and in data analysis with real data in \cref{sec:data:analysis}. Finally we discuss our findings and 
conclude this paper in \cref{sec:conc}.

@@

\iffalse
\subsection{Causal effect adjustment} The DSS method explained
above give us adjusted coefficient estimates for the treatment
and outcome model. However, as result the causal effect estimate
remains the same and modelling with such adjusted sparse effect 
will contribute to the prediction error. Therefore, we need to 
adjust the causal effect estimate as well. We do that in three steps
similar to double residual regression procedure as suggested by
\citet{robinson1988}.

@@

\fi
\subsection{Robust Bayesian Analysis}
We perform our
robust Bayesian analysis on $q\coloneqq(q_1$, \dots, $q_p)\in\mathcal{P}$, where
\begin{equation}
\mathcal{P} \coloneqq \mathcal{P}_1\times\cdots\times\mathcal{P}_p\subseteq \left(0, 1\right)^{p}.
\end{equation}

>>>>

\title{Robust Bayesian Analysis of Causal Inference Problems}
\author{Tathagata Basu$^1$}
\email{tathagatabasumaths@gmail.com}
\author{Matthias C.~M.~Troffaes$^2$}
\email{matthias.troffaes@durham.ac.uk}
\author{Jochen Einbeck$^{2,3}$}
\email{jochen.einbeck@durham.ac.uk}
\address{$^1$Civil and Environmental Engineering, University of Strathclyde}
\address{$^2$Department of Mathematical Sciences, Durham University}
\address{$^3$Durham Research Methods Centre}

\begin{abstract}
Causal inference using observational data is an important aspect in
many fields such as epidemiology, social science, economics, etc. In
particular, our goal is to find the treatment effect on the subjects
along with the causal links between the variables and the outcome.
However, estimation for such problems are extremely 
difficult as the treatment effects may vary from subject 
to subject and modelling the underlying heterogeneity explicitly makes the 
problem practically unsolvable. Another issue we often face is the 
dimensionality of the problem and we need to find a subset of 
explanatory variables to initiate the treatment. However, currently variable selection methods
tend to maximise the predictive performance of the outcome model only. This can be problematic in the case of limited information. 
As the consequence of mistreatment can be harmful. 
So, in this paper, we suggest a general framework with
robust Bayesian analysis which accounts for abstention in deciding
an explanatory variable in the high dimensional regression model. 
To achieve that, we consider a set of spike and slab priors 
through prior elicitation to obtain robust estimates for
both the treatment and outcome model. We are specifically interested 
in the sensitivity of the treatment effect in the high dimensional causal inference
as well as the identifying the confounder variables by means of variable selection. However, indicator
based confounder selection can be deceptive in some cases. Especially, 
when the predictor is strongly associated with either the treatment or 
the outcome. This increases the posterior expectation of the selection
indicators. To avoid that we apply a post-hoc selection scheme
which successfully remove negligible non-zero effects from the model
attaining a smaller set of confounders. Finally, we illustrate
our result using synthetic \alert{and real} dataset.

@@

Bayesian approach in causal effect estimation has become a popular
topic in recent days. However, it has been a topic of interest for about 50 
years and one of the earlier works on this can be
found in \cite{rubin1978}. Lately with the notion of high dimensional problems, Bayesian methodologies have become
more appealing. \citet{Crainiceanu2008} proposed a bi-level 
Bayesian model averaging based method for estimating the causal 
effect. \citet{wang2015} suggested BAC (or, Bayesian adjustment for
confounding) where they use an informative prior obtained from
the treatment model and apply them on the outcome model for
estimating causal effect. Several other methods were also
proposed to tackle confounders from the point of view of Bayesian
variable selection such as: \citet{Zigler2014}, \citet{Hahn2018} etc.

In this paper we take inspiration from the approach of \citet{koch2020}, where
they proposed a bi-level spike and slab prior for causal effect 
estimation. They considered a data-driven adaptive approach to
propose their prior which reduce the variance of the causal estimate. 
In our approach, we are interested in the decision theoretic notion of 
variable selection and its effect in causal estimation. We suggest a 
sensitivity analysis based approach where instead of using a single prior, 
we consider a set of priors \cite{BERGER1990303}. This is particularly 
interesting as in many cases, causal effect estimation can be performed 
through a meta analysis and hence robust Bayesian analysis can be beneficial
\cite{raices_cruz22} under severe uncertainty. Moreover, for some problems 
we have to rely on very limited data to perform our Bayesian analysis and 
choosing a data-driven prior can lead to overfitting. Instead we can 
benefit from expert opinion and elicit prior based on empirical evidence.
This also allows us to construct the problem of confounder selection 
in a framework where abstention has a positive gain. To propose our 
framework, we consider a set of continuous spike and slab priors 
\cite{ishwaran2005} for confounder selection and construct a Bayesian 
group LASSO \cite{xu2015} type problem. To perform the prior sensitivity 
analyses, we consider a set of beta priors on the covariate selection 
probability of the spike and slab priors. We use the posteriors of these
covariate selection probability for identifying the confounders. Finally, 
we consider a post-hoc coefficient adjustment method \citet{hahn2015}
to recover sparse estimates. This post hoc treatment also helps us to 
reduce the bias of treatment effect estimation.

The rest of the paper is organised as follows. In \cref{sec:causal}
we give a formal description of causal estimation problem in the
context of linear regression. \cref{sec:bayes} is focused on the
Bayesian analysis of causal inference problems, followed by the
motivation of a robust Bayesian analysis along with our proposed decision 
theoretic framework for confounder (variable) selection. In \cref{sec:sim}, 
we provide our result of simulation studies under different scenarios 
and data analysis with \alert{real data} in \cref{sec:data:analysis}. Finally 
we discuss our findings and conclude this paper in \cref{sec:conc}.

@@

\iftrue
\subsection{Causal effect adjustment} The DSS method explained
above give us adjusted coefficient estimates for the treatment
and outcome model. However, as result the causal effect estimate
remains the same and modelling with such adjusted sparse effect 
will contribute to the prediction error. Therefore, we need to 
adjust the causal effect estimate as well. We do that in three steps
similar to double residual regression procedure as suggested by
\citet{robinson1988}.

@@

\fi
